---
title: "Association test - Overall samples"
author: "Mun-Gwan Hong"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: 
      collapse : false
    toc_depth: 3
    code_folding: hide
    number_sections: true
    css: styles.css
bibliography: citations_in_report.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding= encoding, output_dir= "../reports") })
---

```{r, echo=FALSE, include=FALSE}
# the default, when no passover
if(!exists("panels")) panels <- c("Inflammation", "Neurology")
cache_path <- paste(c("../cache/rmd/", panels, "-overall-"), collapse = "")

## Chunk hooks
knitr::knit_hooks$set(
  optipng = knitr::hook_optipng   # Ready to activate optipng (Optimize PNG)
)
## Set default chunk options
knitr::opts_chunk$set(
	# dev = c('pdf'), fig.path = '../results/figures/', # export figures to pdf files
  results = "asis",   # for HTML
  message = FALSE,    # no message from the code
  warning = FALSE,    # no warning from the code
	cache.path = cache_path,   # cache the results from heavy analyses
  fig.width  = 8,
  fig.height = 6,
	eval.after = "fig.cap",    # to add fig. caption within the chunk
  optipng = '-o1', # '-o7 -zc9 -zm8 -zs0', # Optimize 'png' plots
	purl = FALSE      # because most of the code are not to be extracted
)
options(knitr.kable.NA = '')  # NA is shown as ''
```

```{r, echo=FALSE, include=FALSE, purl=TRUE}
## Pre-loaded packages
# Please note that some functions are called directly from the package using
# double-colon '::' or triple-colon ':::' to make the source clearer.
library(tidyverse)
library(kableExtra)
library(foreach)  # %dopar%, foreach
```

```{r assoc_res.load_data, purl=TRUE}
## File names
fn <- list(
  lib = list(
    general = "utils.R",
    assoc = "anal-assoc_protein-utils.R"
  ),
  i = list(                               # input
    olk02 = "../data/s1-olink_proteomic.v02.RData",
    c02 = "../data/s1-clinical.v02.RData",
    perm_t = "../data/lm_resample_t.RData"
	),
  o = list(         # intermediate results
    res = paste0("../cache/assoc-overall_samples-", 
                 paste(panels, collapse = ","), ".RData")
  )
)
# Brief check if all files exist
stopifnot(all(file.exists(unlist(fn$lib), unlist(fn$i), dirname(unlist(fn$o)))))

source(fn$lib$general)  # Words, ...
source(fn$lib$assoc)  # functions for association tests

# Load data
load(fn$i$olk02) # proteomic data 
load(fn$i$c02)  # clinical information



####### < SELECTED PANELS > #######
olink <- olink %>% 
  filter(Panel %in% panels)
qns <- qns %>% 
  filter(SampleID %in% olink$SampleID)
# combine `olink` and `qns` for association tests
olink_qns <- left_join(olink, qns, by = "SampleID")



# Load T statistics collected during resampling for max T test
if(!exists("resample_t_res")) load(fn$i$perm_t)
# number of permutation
N_PERMUTATION <- ncol(resample_t_res[["overall"]][[1]])
# Initialize to store association test results
assoc_res <- list()
```

All analyses below were limited to Olink **`r panels`** panel(s).


## Background

### Dimension

The numbers of samples and proteins to analyze are as below.

```{r}
olink_qns %>% 
  group_by(Panel) %>% 
  summarise(
    Samples = n_distinct(SampleID),
    Proteins = n_distinct(OlinkID)
  ) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE)
```

### Prior observation on relevant variables

#### Body-mass-index (BMI)

BMI was **missing** for a relatively large proportion of participants
(N = `r sum(is.na(qns$BMI))`, 
`r round(sum(is.na(qns$BMI)) / dim(qns)[1L] * 100, 1)`%).

#### Smoking

```{r}
# confirm the statement below
stopifnot(all(is.na(qns$BMI[is.na(qns$Smoking)])))
```

Smoking status was **missing** for a relatively large proportion of participants
(N = `r sum(is.na(qns$Smoking))`, 
`r round(sum(is.na(qns$Smoking)) / dim(qns)[1L] * 100, 1)`%).
For all of them, BMI was also missing.

#### Hearing problem

Hearing problem data (`TSCHQ_26`) was **missing** for a relatively large proportion of participants
(N = `r sum(is.na(qns$TSCHQ_26))`, 
`r round(sum(is.na(qns$TSCHQ_26)) / dim(qns)[1L] * 100, 1)`%). 
Most of them were different individuals from those whose BMI was missing. 
Only `r sum(is.na(qns$TSCHQ_26) & is.na(qns$BMI))` were in common.

```{r assoc_res.covariates, purl=TRUE, cache=TRUE}
# Tibble of covariates and association test functions
cvar_test <- tribble(
  ~rhs,                 ~test,
  "Age",                "lm",
  "Sex",                "lm",
  "BMI",                "lm",
  "Smoking",            "aov",
  "`Sample Lab`",       "aov",
  "Age + `Sample Lab`", "aov",
  "`Collection Date`",  "lm",
  "PlateID",            "aov"
) %>% 
  # one line version of those tests
  mutate(f_1line = map(test, ~ get(paste0(.x, "_out_1line"))))

# Association tests
assoc_res$covariates <- map(
  1:nrow(cvar_test),
  ~ olink_assoc_1lines(olink_qns,
                       cvar_test$rhs[.x],
                       cvar_test$f_1line[[.x]])
) %>% 
  setNames(nm = cvar_test$rhs)

# tested covariates
tested_cvars <- unique(map_chr(assoc_res$covariates, ~ .x$Term[1]))
```

### Covariates

`r Words(n_distinct(tested_cvars))` potential covariates were tested for the association with individual protein profiles.
Several proteins were found correlated with some of those covariates (Bonferroni adjusted _p_ value < 0.05).
Especially, quite a large number of proteins were significantly associated with **age, sex, BMI and sample lab**. 
So, those variables were included in most of following analyses as covariates.

Summary table is shown below.
Results in more detail follow. 

```{r}
n_prots <- n_distinct(olink_qns$OlinkID)
# summary table of the association between individual proteins and covariates
assoc_res$covariates %>% 
  map_dfr(., ~ select(.x, Protein, Model, Term, P_bonf)) %>% 
  group_by(Term, Model) %>% 
  summarise(
    prots = list(Protein[P_bonf < 0.05]),
    N = map_int(prots, length),
    Propn. = sprintf("%4.1f%%", N / n_prots * 100),
    Proteins = map_chr(prots, ~ str_trunc(paste(.x, collapse = ", "), 20)),
    .groups = "drop"
  ) %>% 
  select(-prots) %>%
  rename(`Nr. or assoc.\nproteins` = N) %>% 
  mutate(
    # show in the order as `cvar_test`
    Term = fct_relevel(Term, tested_cvars) %>%
      fct_recode("Sex" = "SexFemale"),
    Model = gsub('`', '', Model)
  ) %>% 
  arrange(Term) %>% 
  kbl(align = "llrrl") %>% 
  kable_styling()
```

#### Age

```{r}
# increasing/decreasing proteins as age advances
hits <- filter(assoc_res$covariates$Age, P_bonf < 0.05) %>%
  split(., recode(sign(.$Estimate), "1" = "inc", "-1" = "dec"))
hits$n_all <- nrow(bind_rows(hits))
```

`r Words(hits$n_all)` proteins were found correlated with age by linear regression.
Among them, the profiles of `r nrow(hits$inc)` proteins (`r round(nrow(hits$inc) / hits$n_all * 100, 1)`%) increased as age advances, 

##### Increasing protein as age advances

```{r, fig.height=3}
show_hit_table(hits$inc) %>%
  scroll_box(height = "180px")
show_examples(hits$inc$OlinkID[1:3], Age, olink_qns) +
  geom_point() +
  ggtitle("Top 3 proteins increasing as age advances")
```

##### Decreasing protein as age advances

```{r, fig.height=3}
show_hit_table(hits$dec) %>%
  scroll_box(height = "180px")
show_examples(hits$dec$OlinkID[1:3], Age, olink_qns) +
  geom_point() +
  ggtitle("Top 3 proteins decreasing as age advances")
```

#### Sex

```{r}
# proteins with differential profiles between sexes
hits <- filter(assoc_res$covariates$Sex, P_bonf < 0.05) %>%
  split(., recode(sign(.$Estimate), "1" = "h_f", "-1" = "h_m"))
hits$n_all <- nrow(bind_rows(hits))
```

`r Words(hits$n_all)` proteins were found correlated with sex by linear regression.
Among them, the profiles of `r nrow(hits$h_f)` proteins (`r round(nrow(hits$h_f) / hits$n_all * 100, 1)`%) were higher in females.

##### Higher in females

```{r, fig.height=3}
show_hit_table(hits$h_f) %>%
  scroll_box(height = "180px")
show_examples(hits$h_f$OlinkID[1:3], Sex, olink_qns) +
  geom_boxplot() +
  ggtitle("Top 3 proteins higher in females")
```

##### Higher in males

```{r, fig.height=3}
show_hit_table(hits$h_m) %>%
  scroll_box(height = "180px")
show_examples(hits$h_m$OlinkID[1:3], Sex, olink_qns) +
  geom_boxplot() +
  ggtitle("Top 3 proteins higher in males")
```

#### BMI

```{r}
# increasing/decreasing proteins as BMI
hits <- filter(assoc_res$covariates$BMI, P_bonf < 0.05) %>%
  split(., recode(sign(.$Estimate), "1" = "inc", "-1" = "dec"))
hits$n_all <- nrow(bind_rows(hits))
```

`r Words(hits$n_all)` proteins were found correlated with BMI by linear regression.
Among them, the profiles of `r nrow(hits$inc)` proteins (`r round(nrow(hits$inc) / hits$n_all * 100, 1)`%) increased as BMI, 

##### Increasing protein as BMI

```{r, fig.height=3}
show_hit_table(hits$inc) %>%
  scroll_box(height = "180px")
show_examples(hits$inc$OlinkID[1:3], BMI, olink_qns) +
  geom_point() +
  ggtitle("Top 3 proteins increasing as BMI")
```

##### Decreasing protein as BMI

```{r, fig.height=3}
show_hit_table(hits$dec) %>%
  scroll_box(height = "180px")
show_examples(hits$dec$OlinkID[1:3], BMI, olink_qns) +
  geom_point() +
  ggtitle("Top 3 proteins decreasing as BMI")
```

#### Smoking (`A7`)

```{r}
hits <- filter(assoc_res$covariates$Smoking, P_bonf < 0.05)
```

`r Words(nrow(hits))` proteins were found correlated with smoking status by ANOVA. 

```{r, fig.height=3}
show_hit_table(hits) %>%
  scroll_box(height = "180px")
show_examples(hits$OlinkID[1:3], Smoking, olink_qns) +
  geom_boxplot() +
  ggtitle("Top 3 proteins associated with smoking status") +
  theme(axis.text.x = element_text(angle = 15, margin = margin(t = 10)))
```

#### Sample Lab

```{r}
hits <- map(
  c("`Sample Lab`", "Age + `Sample Lab`") %>% 
    `names<-`(., str_remove_all(., "`")),   # inline access
  ~ filter(assoc_res$covariates[[.x]], P_bonf < 0.05)
)
hits$intx <- hits[["Age + Sample Lab"]] %>% 
  filter(Protein %in% hits[["Sample Lab"]]$Protein)
```

`r Words(nrow(hits[["Sample Lab"]]))` proteins were found different across sample labs by ANOVA. 
In order to take into account dissimilar age distribution between sample labs, the association was also tested after adjusting for age.
`r Words(nrow(hits[["Age + Sample Lab"]]))` proteins were found after the adjustment.
Among them, `r nrow(hits$intx)` proteins were in common, the results of which for the adjusted model are listed below.

```{r, fig.height=3}
show_hit_table(hits$intx) %>%
  scroll_box(height = "180px")
show_examples(hits$intx$OlinkID[1:3], `Sample Lab`, olink_qns) +
  geom_boxplot() +
  ggtitle("Top 3 proteins different across sample labs")
```

#### Collection Date (age of sample)

```{r}
# increasing/decreasing proteins as the collection date
hits <- filter(assoc_res$covariates[["`Collection Date`"]], P_bonf < 0.05)
```

`r Words(nrow(hits))` protein was found correlated with sample collection date or the age of samples by linear regression.

```{r, fig.width=4, fig.height=3}
show_hit_table(hits)
show_examples(hits$OlinkID, `Collection Date`, olink_qns) +
  geom_point()
```

#### Assay Plate

```{r}
hits <- filter(assoc_res$covariates$PlateID, P_bonf < 0.05) 
```

`r Words(nrow(hits))` protein was associated with assay plate. 

------------------------

## Association test for individual proteins

### Notes for the result tables in following sections

- The 'p value' is the nominal p value without any multiple correction.

- The **'Adj.P by perm.'** is the adjusted P value by $`r N_PERMUTATION`$ permutation 
using Westfall and Young's max-T method ([Westfall and Young, 1993](https://www.google.com/books?hl=en&lr=&id=nuQXORVGI1QC&oi=fnd&pg=PR11&dq=resampling-based+multiple+testing+examples+and+methods+for+p-value+adjustment+pdf&ots=XngUGMa2LO&sig=Vr87y-1gzC_Glhz3sB6N9R95MU0)). 
The resampling-based procedure can take into account correlation structure in data.
It is often considered as a good method to handle multiple testing correction, but it is computationally heavy. 

- The 'q value' was computed using 'qvalue' (v `r packageVersion("qvalue")`) R package ([Ref](https://projecteuclid.org/euclid.aos/1074290335)) to adjust the p values for the number of hypothesis tests. 

- Those **multiple testing corrections** (given as 'Adj.P by perm.' and 'q value') were conducted **per table**. 
In other words, it was for all the tests of **a selected model only**. 
Those corrections were not applied for controlling the family-wide error rate or positive false discovery rate of overall tests in this report. 
As an example, the q values from the models for Tinnitus status including age, sex and smoking as covariate didn't take into account the tests without any adjustment nor the tests with hearing problem. 


### Assoc. with tinnitus status

#### Notes

##### Data source

- The tinnitus status was obtained from LifeGene questionnaire table, in which it was labeled as `Intro_3`.

##### Statistical models for association tests

- Linear regression model for the association between one protein and Tinnitus status, including a varying set of covariates for adjustment of the effects of them. 
The adjustment factors are listed for each model. 

- `lm(Protein ~ (tinnitus status) + covariate1 + covariate2 + ...)`

##### Effect size shown in result tables

- The effect size, labeled as $\Delta$'(Yes - No)' in the following tables, is the difference between tinnitus cases and controls of estimated protein values after adjustment for selected covariates. 
Positive value indicates the estimated value was higher in cases than controls.
Please note that absolute magnitudes are not comparable between proteins, because the NPX values from Olink assays were given in arbitrary units.

#### Main results

```{r}
# No missing in those variables
stopifnot(
  all(!is.na(qns$Age)),
  all(!is.na(qns$Sex)),
  all(!is.na(qns$Tinnitus)),
  all(!is.na(qns$`Sample Lab`))
)
```

```{r assoc_res.perm_t, purl=TRUE, cache=TRUE}
# Association tests
assoc_res$perm_t <- resample_t_res$overall %>% 
  map(~ lm_prot_padj_by_max_t(.x, olink, qns))
```

##### **Without** any adjustment

```{r}
out_df <- assoc_res$perm_t[["Tinnitus"]]
```

* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 

```{r}
show_delta_P_perm(out_df)
rhs <- "Tinnitus + Age + Sex + `Sample Lab`"
out_df <- assoc_res$perm_t[[rhs]]
```

##### Covariates included : `r text_Rmd_terms(rhs)`

* `r text_R_fun_prot(rhs)`
* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 

```{r}
show_delta_P_perm(out_df)
rhs <- "Tinnitus + Age + Sex + BMI"
out_df <- assoc_res$perm_t[[rhs]]
```

##### Covariates included : `r text_Rmd_terms(rhs)`

* `r text_R_fun_prot(rhs)`
* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 
* Please note that the number of samples included in this analysis was substantially lower, because BMI of `r round(sum(is.na(qns$BMI)) / nrow(qns) * 100, 1)`% of samples was missing.

```{r}
show_delta_P_perm(out_df)
rhs <- "Tinnitus + Age + Sex + Smoking"
out_df <- assoc_res$perm_t[[rhs]]
```

##### Covariates included : `r text_Rmd_terms(rhs)`

* `r text_R_fun_prot(rhs)`
* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 
* Please note that the number of samples included in this analysis was substantially lower, because smoking status of `r round(sum(is.na(qns$Smoking)) / nrow(qns) * 100, 1)`% of samples was missing.

```{r}
show_delta_P_perm(out_df)
rhs <- "Tinnitus + Age + Sex + BMI + `Sample Lab` + Smoking"
out_df <- assoc_res$perm_t[[rhs]]
```

##### Covariates included : `r text_Rmd_terms(rhs)`

* `r text_R_fun_prot(rhs)`
* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 
* Please note that the number of samples was lower due to missing smoking status.

```{r}
show_delta_P_perm(out_df)
rhs <- "Tinnitus + Age + Sex + BMI + Smoking + TSCHQ_26"
out_df <- assoc_res$perm_t[[rhs]]
```

##### Covariates included : `r text_Rmd_terms(rhs)`

* `r text_R_fun_prot(rhs)`
* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 

```{r}
show_delta_P_perm(out_df)
rhs <- "Tinnitus + Age + Sex + BMI + `Sample Lab` + Smoking + TSCHQ_26"
out_df <- assoc_res$perm_t[[rhs]]
```

##### Covariates included : `r text_Rmd_terms(rhs)`

* `r text_R_fun_prot(rhs)`
* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 
* Due to missing BMI, smoking and/or hearing problem status, only `r nrow(drop_na(qns, BMI, Smoking, TSCHQ_26))` samples were included in this analysis.

```{r}
show_delta_P_perm(out_df)
```

#### Limited to non-hearing aid users

```{r assoc_res.non_aid_user, purl=TRUE, cache=TRUE}
qns_non_aids <- filter(qns, non_aids_analysis)
tmp <- filter(olink, SampleID %in% qns_non_aids$SampleID)
assoc_res$non_aid_user <- map(
  resample_t_res$non_aid_user,
  ~ lm_prot_padj_by_max_t(.x, tmp, qns_non_aids)
)
```

* There were `r sum(qns$A14_1 == "Yes", na.rm = TRUE)` hearing aid users.
After excluding the users together with matched non-users, `r nrow(qns_non_aids)` were remained for this analysis.

```{r}
rhs <- "Tinnitus + Age + Sex + `Sample Lab`"
out_df <- assoc_res$non_aid_user[[rhs]]
```

##### Covariates included : `r text_Rmd_terms(rhs)`

* `r text_R_fun_prot(rhs)`
* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 

```{r}
show_delta_P_perm(out_df)
rhs <- "Tinnitus + Age + Sex + BMI + `Sample Lab` + Smoking + TSCHQ_26"
out_df <- assoc_res$non_aid_user[[rhs]]
```

##### Covariates included : `r text_Rmd_terms(rhs)`

* `r text_R_fun_prot(rhs)`
* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 
* Due to missing BMI, smoking and/or hearing problem status, only `r nrow(drop_na(qns_non_aids, BMI, Smoking, TSCHQ_26))` samples were included in this analysis.

```{r}
show_delta_P_perm(out_df)
```

#### Additional results with adjustment for stress variables

```{r}
ss <- qsets$stress
rhs1 <- "Tinnitus + Age + Sex + BMI + Smoking + `Sample Lab`"
```

- Stress related variables : `r knitr::combine_words(ss$with_q)`.

- Tests for association between tinnitus and individual proteins

- Due to missing information, `r nrow(drop_na(qns, BMI, Smoking, all_of(ss$lab)))` samples were analyzed here.

##### Individual stress related variables

```{r}
rhss <- map_chr(1:nrow(ss), ~ paste(rhs1, "+", ss$with_q[.x]))
is_hit <- map(assoc_res$perm_t[rhss], ~ (.x$P_perm < 0.05))
rhs_x <- paste(rhs1, "+ `stress i`")
```

* Covariates included : **individual stress related variables** as well as `r text_Rmd_terms(rhs1)`
* `r text_R_fun_prot(rhs_x)`
* `r Words(sum(unlist(is_hit)))` protein was significantly associated with one of stress variables (adjusted P by permutation < 0.05). 

```{r}
for(ii in 1:nrow(ss)) {
  print(show_delta_P_perm(
    assoc_res$perm_t[[rhss[ii]]], 
    caption = paste0(ss$desc[ii], " (", ss$lab[ii], ") - Top 5")
  ))
}
```

##### **All `r words(nrow(ss))` stress related variables**

```{r}
rhs <- paste(rhs1, "+", paste(ss$with_q, collapse = " + "))
out_df <- assoc_res$perm_t[[rhs]]
txt <- paste("All", nrow(ss), "stress related variables")
rhs_x <- paste(rhs1, "+ `stress 1` + `stress 2` + `...` + `stress i`")
```

* Covariates included : **`r txt`** as well as `r text_Rmd_terms(rhs1)`
* `r text_R_fun_prot(rhs_x)`
* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 

```{r}
show_delta_P_perm(out_df, caption = paste(txt, "- Top 5"))
```

----------------------------

### Assoc. with stress

#### Notes

##### Data source

- Stress related variables

```{r}
ss <- qsets$stress
select(ss, "Variable ID" = "lab", "Description" = "desc") %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE)
# covariates
rhs1 <- "Tinnitus + Age + Sex + BMI + Smoking + `Sample Lab`"
```

- Due to missing information, `r nrow(drop_na(qns, BMI, Smoking, all_of(ss$lab)))` samples were analyzed here.

##### Statistical models for association tests

- Linear regression model or ANOVA for the association between one protein and **individual stress related variables**, adjusting for `r text_Rmd_terms(rhs1, 0)`
* `r text_R_fun_prot(paste("\x60stress i\x60 +", rhs1))`

```{r}
rhss <- map_chr(1:nrow(ss), ~ paste(ss$with_q[.x], "+", rhs1))
```

##### Effect size shown in result tables

- The effect size, labeled as 'Estimate' in the following tables, is the change by one unit increment of the scores (e.g. from 0 to 1) or $\Delta$(`No` - `Yes`), the difference between `No` and `Yes`.
- Please note that absolute magnitudes are not comparable between proteins, because the NPX values from Olink assays were given in arbitrary units.

#### Results

```{r}
is_hit <- map(assoc_res$perm_t[rhss], ~ (.x$P_perm < 0.05))
```

* `r Words(sum(unlist(is_hit)))` protein was significantly associated with one of stress variables (adjusted P by permutation < 0.05). 

```{r, fig.width=5, fig.height=4}
for(ii in 1:nrow(ss)) {
  p <- assoc_res$perm_t[[rhss[ii]]] %>% 
    mutate(qval = qvalue::qvalue(Pval)$qvalues) %>% 
    show_res_p(
      Protein, Estimate, "Adj.P by perm." = P_perm, "_q_ value" = qval,
      caption = paste0(ss$desc[ii], " (", ss$lab[ii], ")")
    )
  print(p)
} 

for(ii in 1:nrow(ss)) {
  if(any(is_hit[[ii]])) {
    hits <- assoc_res$perm_t[[rhss[ii]]]$OlinkID[is_hit[[ii]]]
    p <- olink_qns %>% 
      filter(OlinkID %in% hits) %>%
      drop_na(NPX, Protein, all_of(ss$lab[ii])) %>% 
      ggplot() +
      aes_string(x = ss$with_q[ii], y = "NPX") +
      facet_wrap(~ Protein) +
      ggtitle(paste0(ss$desc[ii], " (", ss$lab[ii], ")"))
    if(n_distinct(olink_qns[[ss$lab[ii]]]) == 2) {
      print(p + geom_boxplot() + add_n())
    } else {
      print(p + geom_point())
    }
  }
}
```

----------------------------

### Assoc. with hearing problem (`TSCHQ_26`)

#### Notes

##### Statistical models for association tests

```{r}
rhs <- "TSCHQ_26 + Age + Sex + BMI + Smoking + `Sample Lab`"
out_df <- assoc_res$perm_t[[rhs]]
```

- Linear regression model for the association between one protein and **hearing problem (Yes vs. No)**, adjusting for `r text_Rmd_terms(rhs)`
* `r text_R_fun_prot(rhs)`

#### Results

* `r Words(sum(out_df$P_perm < 0.05))` protein was significantly associated with tinnitus (adjusted P by permutation < 0.05). 

```{r}
out_df %>% 
  mutate(qval = qvalue::qvalue(Pval)$qvalues) %>% 
  show_res_p(Protein, 
             "$\\Delta$(No - Yes)" = Estimate, 
             "Adj.P by perm." = P_perm, 
             "_q_ value" = qval)
```

----------------------------

### Assoc. with tinnitus subtypes (`TSCHQ_??`)

```{r}
styp <- qsets$tinnitus_subtypes
```

#### Categorical variables

- `r Words(length(styp$cat))` categorical variables were tested (`r knitr::combine_words(styp$cat, before = '\x60')`).

##### Statistical models for association tests

```{r}
rhs1 <- "Age + Sex + BMI + Smoking + `Sample Lab`"
```

- ANOVA for the association between one protein and **individual sub-type variables**, including `r text_Rmd_terms(rhs1, 0)`
- `r text_R_fun_prot(paste(rhs1, "+ subtype_i"), "aov", -1)`
- Please note the number of samples analyzed here was substantially lower, due to missing data of many samples.

##### Homoscedasticity assumption of ANOVA

Because it is known that ANOVA is sensitive to **homoscedasticity** assumption (homogeneity of variance),
_p_ values from Bartlett's test were shown in the result tables. 
When the _p_ value is significant, the association detected by ANOVA was shown significant possibly by heteroscedasticity rather than real difference in mean. 

```{r assoc_res.subtype.cat, purl=TRUE, cache=TRUE}
# Association tests
assoc_res$subtype$cat <- map(
  setNames(nm = styp$cat),
  ~ olink_assoc_1lines(olink_qns,
                       paste(rhs1, "+", .x),
                       aov_bartlett_out_1line)
)
```

##### Results

```{r}
out_df <- assoc_res$subtype$cat
is_hit <- map(out_df, ~ {.x$qval < 0.05})
```

* `r Words(sum(unlist(is_hit)))` proteins were significantly associated with one of those categorical sub-type variables (_q_ value < 0.05).

```{r}
all_t_rhs1 <- all.vars(as.formula(paste("~", rhs1)))
for(iv in styp$cat) {
  if(any(is_hit[[iv]])) {
    qn <- paste(iv, ":", codes$Label[codes$Variable == iv])
    
    hits <- out_df[[iv]][is_hit[[iv]], ]
    print(show_res_p(hits, Protein, "_q_ value" = qval, 
                     caption = qn, n = Inf, after_p = "Bartlett P"))

    p <- olink_qns %>% 
      filter(OlinkID %in% hits$OlinkID) %>%
      drop_na(NPX, all_of(c(all_t_rhs1, iv))) %>% 
      ggplot() +
      aes_string(x = iv, y = "NPX") +
      facet_wrap(~ Assay, ncol = 2, scales = "free_y") +
      geom_boxplot() +
      add_n() +
      theme(axis.text.x = element_text(angle = 10, margin = margin(t = 10))) +
      ggtitle(qn)
    print(p)
  }
}
```

#### Continuous or two-group variables

- `r Words(length(styp$qty))` continuous variables were tested (`r knitr::combine_words(styp$qty, before = '\x60')`).

##### Statistical models for association tests

```{r}
rhs1 <- "Age + Sex + BMI + Smoking + `Sample Lab`"
```

- Linear regression for the association between one protein and **individual sub-type variables**, including `r text_Rmd_terms(rhs1, 0)`
- `r text_R_fun_prot(paste("subtype_i +", rhs1))`

```{r assoc_res.subtype.qty, purl=TRUE, cache=TRUE}
# Association tests
assoc_res$subtype$qty <- map(
  setNames(nm = styp$qty),
  ~ olink_assoc_1lines(olink_qns,
                       paste(iv, "+", rhs1),
                       lm_out_1line)
)
```

##### Results

```{r}
out_df <- assoc_res$subtype$qty
is_hit <- map(out_df, ~ {.x$qval < 0.05})
```

* `r Words(sum(unlist(is_hit)))` protein was significantly associated with one of those continuous sub-type variables (_q_ value < 0.05).

```{r}
all_t_rhs1 <- all.vars(as.formula(paste("~", rhs1)))
for(iv in styp$qty) {
  if(any(is_hit[[iv]])) {
    qn <- paste(iv, ":", codes$Label[codes$Variable == iv])

    hits <- out_df[[iv]][is_hit[[iv]], ]
    print(show_res_p(hits, Protein, "Change by 1 in score" = Estimate, 
                     "q value" = qval, caption = qn, n = Inf))
    p <- olink_qns %>% 
      filter(OlinkID %in% hits$OlinkID) %>%
      drop_na(NPX, all_of(c(all_t_rhs1, iv))) %>% 
      ggplot() +
      aes_string(x = iv, y = "NPX") +
      facet_wrap(~ Assay, ncol = 2) +
      geom_point() +
      ggtitle(qn)
    print(p)
  }
}
```

----------------------------

### Tinnitus Handicap Inventory `THI_??`

#### Individual THI scores

##### Data source and statistical models

```{r}
scr <- qsets$THI_scores
# Confirm every THI score is ordered
stopifnot(all(map_lgl(scr, ~ is.ordered(qns[[.x]]))))
# Confirm the number of included samples is constant
tmp <- drop_na(qns, Smoking)
tmp <- map_int(scr, ~ sum(!is.na(tmp[[.x]])))
stopifnot(all(tmp == tmp[1]))
# descriptive RHS
rhs1 <- "Age + Sex + BMI + Smoking + `Sample Lab`"
rhs_x <- paste("`a THI score` +", rhs1)
```

- There were `r length(scr)` tinnitus handicap inventory (THI) variables, `THI_??`.
- Every THI score was coded as ordered categories (e.g. `Yes`, `Sometimes` and `No`).
- Linear regression for the association between one protein and **individual THI scores**, including `r text_Rmd_terms(rhs_x)`
- `r text_R_fun_prot(rhs_x)`
- The _p_ values were adjusted for all tested proteins and all `r length(scr)` THI variables.
- Due to missing information, `r nrow(drop_na(qns, Smoking, all_of(scr)))` samples were analyzed here.

```{r assoc_res.THI_scores, purl=TRUE, cache=TRUE}
# Association tests
assoc_res$THI_scores <- lapply(
  scr, 
  function(ii) {
    f <- formula(paste("NPX ~", ii, "+", rhs1))
    olink_qns %>%
      nest(data = -c(OlinkID, Assay, Panel)) %>%
      add_column(map_dfr(.$data, ~ lm_out_1line(f, .x))) %>%
      select(-data) %>% 
      rename(Protein = Assay)
  }
) %>% 
  bind_rows() %>%    # all into one 
  mutate(qval = qvalue::qvalue(Pval)$qvalues)
```

##### Results

```{r}
out_df <- assoc_res$THI_scores %>% 
  mutate(Term = sub(".L", "", Term))
is_hit <- out_df$qval < 0.05
```

* `r Words(sum(is_hit))` protein was significantly associated with one of those THI scores (_q_ value < 0.05).

```{r, fig.height=4, fig.width=6}
if(any(is_hit)) {
  hits <- out_df[is_hit, ]
  
  print(show_res_p(
    hits, Protein, Term, "Trend Yes-Sometimes-No" = Estimate,
    "_q_ value" = qval, n = Inf, caption = "THI scores"
  ))

  for(ii in unique(hits$Term)) {
    p <- olink_qns %>% 
      filter(OlinkID %in% hits$OlinkID) %>%
      # sort by Pval
      mutate(Protein = factor(Assay, levels = hits$Protein)) %>% 
      drop_na(NPX, Protein, Smoking, all_of(ii)) %>% 
      ggplot() +
      aes_string(x = ii, y = "NPX") +
      facet_wrap(~ Protein, ncol = 3, scales = "free_y") +
      geom_boxplot() +
      add_n() +
      ggtitle(paste(ii, ":", codes$Label[codes$Variable == ii]))
    print(p)
  }
} else {
  print(show_res_p(
    out_df, Protein, Term, "Trend Yes-Sometimes-No" = Estimate,
    "_q_ value" = qval, n = 3, caption = "THI scores - Top 3"
  ))
}
```

#### THI total score

```{r}
rhs <- "`THI Total score` + Age + Sex + BMI + Smoking + `Sample Lab`"
```

- Linear regression for the association between one protein and **THI total score**, 
including `r text_Rmd_terms(rhs)`
- `r text_R_fun_prot(rhs)`
- Due to missing information, `r nrow(drop_na(qns, Smoking, all_of("THI Total score")))` samples were analyzed here.

##### Numeric score

```{r assoc_res.THI_total_score.num, purl=TRUE, cache=TRUE}
# Association tests
out_df <- assoc_res$THI_total_score$num <- 
  olink_assoc_1lines(olink_qns, rhs, lm_out_1line)
```

* `r Words(sum(out_df$qval < 0.05))` protein was significantly associated with THI total score as numeric (_q_ value < 0.05).

```{r}
show_res_p(out_df, Protein, "Trend" = Estimate, "_q_ value" = qval)
```

##### Ordinal categorical scores

```{r}
score_cut <- tribble(
  ~up_bound, ~lab,
   16,       "Negligible", 
   36,       "Light", 
   56,       "Moderate", 
   76,       "Severe", 
  100,       "Catastrophic"
)

# Explanation on the categorization
score_cut %>% 
  mutate(Range = paste(lag(up_bound, default = -1) + 1, "-", up_bound)) %>% 
  select(Label = lab, Range) %>% 
  kbl(align = "lc") %>% 
  kable_styling(full_width = FALSE)
```

```{r assoc_res.THI_total_score.ord, purl=TRUE, cache=TRUE}
# Association tests
out_df <- assoc_res$THI_total_score$ord <- olink_qns %>% 
  mutate(
    # ordered categories
    THI.Total.score.cat = cut(
      `THI Total score`, 
      breaks = c(-1, score_cut$up_bound),
      labels = score_cut$lab,
      ordered_result = T
    )
  ) %>%
  olink_assoc_1lines(
    "THI.Total.score.cat + Age + Sex + BMI + Smoking + `Sample Lab`",
    lm_out_1line
  )
```

* `r Words(sum(out_df$qval < 0.05))` protein was significantly associated with THI total score as ordered categorical (_q_ value < 0.05).

```{r}
show_res_p(out_df, Protein, "Trend" = Estimate, "_q_ value" = qval)
```

------------------------

### Other selected variables

#### Statistical test models

```{r}
others <- tribble(
  ~lab,      ~test,  ~desc,
  "Intro_8",  "lm", "Gross income",
  "Intro_9", "aov", "Education",
  "TSCHQ_28", "lm", "Tolerate loud sounds",
  "A15_4",    "lm", "Temporomandibular joint pain",
  "A12",      "lm", "Solely sensitive"
) %>% 
  # one line version of those tests
  mutate(f_1line = map(test, ~ get(paste0(.x, "_out_1line"))))

# descriptive RHS
rhs1 <- "Age + Sex + BMI + Smoking + `Sample Lab` + Tinnitus"
rhs_x <- paste("`a variable` +", rhs1)
```

- Linear regression for the association between one protein and **individual THI scores**, including 
If both cases and controls were included, tinnitus status was also included as an adjusting factor.
- `r text_R_fun_prot(rhs_x)`

```{r assoc_res.others, purl=TRUE, cache=TRUE}
# Association tests
assoc_res$others <- lapply(
  1:nrow(others),
  function(ii) {
    f_rhs <- as.formula(paste("~", others$lab[ii], "+", rhs1))
    no_na <- olink_qns %>%
      drop_na(NPX, all_of(all.vars(f_rhs)))
    
    if(n_distinct(no_na$Tinnitus) < 2) {
      f_rhs <- update.formula(f_rhs, ~ . - Tinnitus)
    }
    rhs <- as.character(f_rhs)[2]
    olink_assoc_1lines(no_na, rhs, others$f_1line[[ii]])
  }
)
```

#### Results

```{r}
out_df <- assoc_res$others
for(ii in 1:nrow(others)) {
  cat(
    "#####", others$desc[ii], ":", paste0("`", others$lab[ii], "`"), "\n",
    "* Question :", codes$Label[codes$Variable == others$lab[ii]], "\n",
    "*", Words(sum(out_df[[ii]]$qval < 0.05)), 
    "protein was significantly associated (_q_ value < 0.05).\n"
  )
  p <- if(others$test[ii] == "lm") {
    out_df[[ii]] %>% 
      show_res_p(Protein, Estimate, "_q_ value" = qval)
  } else {
    out_df[[ii]] %>% 
      show_res_p(Protein, "_q_ value" = qval)
  }
  print(p)
  cat("\n\n")
}
```

```{r, purl=TRUE, include=FALSE}
save(assoc_res, file = fn$o$res)
```

----------------------------

## Methods

_short text about the methods. Please rephrase for manuscripts_

Bioinformatic data handling and statistic analyses were conducted on `r R.version.string`, 
together with tidyverse (v. `r packageVersion("tidyverse")`) package. 
The association between a protein and a clinical trait was tested using linear regression or ANOVA including described covariates. 
Two methods were applied for multiple testing correction, Westfall and Young's max-T method 
[@westfall1993resampling] and _q_ value [@storey2003positive].
Resampling of the former method was conducted $`r N_PERMUTATION`$ times.
The 'q value' was computed using 'qvalue' (v `r packageVersion("qvalue")`) R package. 
Homoscedasticity assumption was checked by Bartlett's test.

