---
params:
  project_title: "Blood biomarkers in chronic tinnitus"
  requester_name: "Christopher Cederroth"
  requester_mail: "christopher.cederroth@ki.se"
  pi_name: "Christopher Cederroth"
  pi_mail: "christopher.cederroth@ki.se"
  organisation: "Karolinska Institutet"
  staff_name: "Mun-Gwan Hong"
  staff_mail: "mungwan.hong@nbis.se"
  staff_web: "https://nbis.se/about/staff/mungwan-hong/"
title: "`r params$project_title`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: 
      collapse : false
    toc_depth: 4
    code_folding: hide
    number_sections: true
    css: styles.css
bibliography: citations_in_report.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding= encoding, output_dir= "../reports") })
---

```{r setup, echo=FALSE, include=FALSE}
## Pre-loaded packages
# Please note that many functions were called directly from the package using 
# double colon '::' to make the source clearer. 
library(tidyverse)
library(kableExtra)

knitr::knit_hooks$set(
  optipng = knitr::hook_optipng   # Optimize PNG plots
)
## Set default chunk options
knitr::opts_chunk$set(
	# dev = c('pdf'), fig.path = '../results/figures/', # export figures to pdf files
  results = "asis",   # for HTML
  message = FALSE,    # no message from the code
  warning = FALSE,    # no warning from the code
  fig.width  = 8,
  fig.height = 6,
	eval.after = "fig.cap",    # to add fig. caption within the chunk
  optipng = '-o1', # '-o7 -zc9 -zm8 -zs0', # Optimize 'png' plots
	purl = FALSE      # because most of the code are not to be extracted
)
options(knitr.kable.NA = '')  # NA is shown as ''
```

# Support project information

 * NBIS staff: **`r params$staff_name`**
 * Request by: **`r params$requester_name`**
 * Principal investigator: **`r params$pi_name`**
 * Organisation: **`r params$organisation`**


# Work log

*A brief summary*

 * **2021-04-30:** First meeting


# Practical information

```{r Report variables, echo = FALSE}
affil <- paste("National Bioinformatics Infrastructure Sweden, Science for",
               "Life Laboratory, Stockholm University")
link_support <- "http://nbis.se/support/support.html"
link_uppmax <- paste("http://uppmax.uu.se/support-sv/faq/general-",
                     "miscellaneous-faq/acknowledging-uppmax--snic--and-uppnex")
link_authorship <- "http://www.icmje.org/recommendations/"
link_agreement <- "http://nbis.se/assets/doc/nbis-support-useragreement.pdf"
link_acknowledge <- paste("https://ngisweden.scilifelab.se/info/faq#how-do",
                          "-i-acknowledge-ngi-in-my-publication")
link_sensitive_data <- "https://nbis.se/support/human-data.html"
```


## Data responsibilities

Unfortunately, NBIS does not have resources to keep any files associated with
the support request; we kindly suggest that you safely store the results
delivered by us. In addition, we kindly ask that you remove the files from
UPPMAX/UPPNEX. The main storage at UPPNEX is optimized for high-speed and
parallel access, which makes it expensive and not the right place for long-term
archiving. Please be considerate of your fellow researchers by not taking up
this expensive space.

The responsibility for data archiving lies with universities and we recommend
asking your local IT for support with long-term data storage. The
[Data Center](https://www.scilifelab.se/data/) at SciLifeLab may also be
of help with discussing other options.

Please note that special considerations may apply to human-derived, sensitive
personal data. This should be handled according to specific laws and
regulations as outlined at the [NBIS website](`r link_sensitive_data`).

## Acknowledgements

If you are presenting the results in a paper, at a workshop or at a conference,
we kindly remind you to acknowledge us according to the signed
[NBIS User Agreement](`r link_agreement`):

[NBIS staff should be included as co-authors](`r params$staff_web`) if the
support work leads to a publication and when this is merited in accordance to
the ethical recommendations for authorship, *i.e.* the [ICMJE
recommendations](`r link_authorship`). If applicable, please include
*`r params$staff_name`, `r affil`* as co-author. If the above is not
applicable, please acknowledge NBIS like so: *Support by NBIS (National
Bioinformatics Infrastructure Sweden) is gratefully acknowledged.*

## Closing procedures

You should soon be contacted by one of our managers, Jessica Lindvall
*(jessica.lindvall@nbis.se)* or Henrik Lantz
*(henrik.lantz@nbis.se)*, with a request to close down the project in our
internal system and for invoicing matters. If we do not hear from you within
**30 days** the project will be automatically closed and invoice sent.
Again, we would like to remind you about data responsibility and
acknowledgements, see the sections on data responsibilities and
acknowledgments.

You are naturally more than welcome to come back to us with further data
analysis request at any time via [the support form](`r link_support`). Thank
you for using NBIS, we wish you the best of luck with your future research!

<br><br>

----------------------------

# Background

## Study design

Blood samples were collected in the LifeGene project, before the participant selection of the STOP project for Tinnitus. 
Patients (chosen on the basis of their answers to the questionaire for LifeGene) were invited to join the STOP study. 
Healthy controls were invited to the study afterwards. 
All of them answered a separate questionaire specially designed for Tinnitus. 
Some healthy controls were identified to have Tinnitus by the Tinnitus questionaire, who had been classified as healthy by the previous LifeGene questionaire. 
Here 548 patients and 548 controls matched for age and sex were analyzed. 

The website for this study : [stop.ki.se](http://stop.ki.se)

# Whole samples in the study

```{r child = "report-sample_info.Rmd"}
```

# QC of clinical data table

```{r child = "report-QC-clinical_info.Rmd"}
```

# QC of Olink Data

```{r child = "report-QC-Olink_proteomic_data.Rmd"}
```

# Overview of the data after QC

```{r child = "report-overview_after_QC.Rmd"}
```

# Association of individual proteins - overall samples

```{r child = "report-assoc-overall_samples.Rmd"}
```

# Association of individual proteins - females only
```{r, include=FALSE}
sex <- "female"
```

```{r child = "report-assoc-by_sex.Rmd"}
```

# Association of individual proteins - males only
```{r, include=FALSE}
sex <- "male"
```

```{r child = "report-assoc-by_sex.Rmd"}
```

----------------------------

# Appendix

## Methods

_short texts about the methods applied globally_

Bioinformatic data handling and statistic analyses were done on `r R.version.string`, 
together with tidyverse (v. `r packageVersion("tidyverse")`) package.

## R environment

```{r sessionInfo, results='markup'}
sessionInfo()
```


