---
title: "Tinnitus - Overview of the data after QC"
author: "Mun-Gwan Hong"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: 
      collapse : false
    toc_depth: 3
    code_folding: hide
    number_sections: true
    css: styles.css
# bibliography: citations_in_report.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding= encoding, output_dir= "../reports") })
---

```{r after_qc - setup, echo=FALSE, include=FALSE}
## Pre-loaded package
# Please note that some functions are called directly from the package using
# double-colon '::' or triple-colon ':::' to make the source clearer.
library(tidyverse)
library(kableExtra)
library(janitor)

rm(list = ls())

## Chunk hooks
knitr::knit_hooks$set(
  optipng = knitr::hook_optipng   # Optimize PNG plots
)
## Set default chunk options
knitr::opts_chunk$set(
	# echo = FALSE,    # comment out for code_folding
	# dev = c('pdf'), fig.path = '../results/figures/', # export figures to pdf files
  results = "asis",   # for HTML
  message = FALSE,    # no message from the code
  warning = FALSE,    # no warning from the code
	cache.path = "../cache/",   # cache the results from heavy analyses
  fig.width  = 8,
  fig.height = 6,
	eval.after = "fig.cap",    # to add fig. caption within the chunk
  optipng = '-o1', # '-o7 -zc9 -zm8 -zs0', # Optimize 'png' plots
	purl = FALSE      # because most of the code are not to be extracted
)
options(knitr.kable.NA = '')  # NA is shown as ''
```

```{r, purl=TRUE}
## File names
fn <- list(
  lib = "utils.R",
	i = list(                               # input
    olk02 = "../data/s1-olink_proteomic.v02.RData", 
    c01 = "../data/s1-clinical.v01.RData"
	)
)

# Brief check if all files exist
stopifnot(all(file.exists(fn$lib, unlist(fn$i))))
source(fn$lib)  # geom_split_violin, add_n

# Load proteomic data ver.02
load(fn$i$olk02)
# Load clinical information ver.01
load(fn$i$c01)

# confirm the clinical information of all samples included in `olink` is
# available in `qns`
stopifnot(all(unique(olink$SampleID) %in% qns$RID))

# Limit to the 1088 samples with proteomic data
qns <- qns %>% 
  filter(RID %in% unique(olink$SampleID))

panels <- unique(olink$Panel)
```

## Summary on QC

### Clinical data

These variables were excluded.

* `r length(about_table$mostly_na)` variables having too many missing values (>80%)
* `r nrow(about_table$one_value_only)` variables that had only one identical value for all samples
* `r nrow(about_table$one_value_or_na)` variables that had only one identical value or missing for all samples

The codes of these variables were fixed.

* `r length(about_table$wrong_code$noyes)` variables for Yes / No question
* `r length(about_table$wrong_code$noyes_little_to_lot)` variables for 'No', 'Yes, a little', 'Yes, quite a lot', 'Yes, a lot' question

### Proteomic data

* `r n_distinct(during_qc$missing_tbl$SampleID)` samples were found to contain missing values. 
Among them, the data of `r n_distinct(unlist(during_qc$i_all_missing))` samples was missing for whole proteins of one panel or two.
* The data of `r unique(during_qc$samples_too_many_llod$Panel)` panel of `r nrow(during_qc$samples_too_many_llod)` sample was excluded, 
because majority (>40%) of them were below LLOD
(`r sprintf("%2.1f%%", during_qc$samples_too_many_llod$below_lod_prop * 100)`).
* The >50% values of `r length(during_qc$i_assys_too_many_llod)` proteins were below LLOD, the data of which were removed.

### Dimension after the QC

The numbers of samples and proteins after the QC were as below.

```{r}
olink %>% 
  group_by(Panel) %>% 
  summarise(
    Samples = n_distinct(SampleID),
    Proteins = n_distinct(OlinkID)
  ) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE)
```

-----------------------------

## Samples with proteomic data

Note, all investigations in this chapter **were limited to** those of the samples of which proteomic data were acquired (N = `r nrow(qns)`). 

### Matching between cases and controls

```{r, purl=TRUE}
# a list of sample IDs divided by panel
sample_id_by_panel <- split(olink$SampleID, olink$Panel) %>% 
  lapply(unique)
```

#### Tinnitus status and Sex

There were two groups in terms of disease status of tinnitus, `r knitr::combine_words(unique(qns$Tinnitus), before = "'")`. 
These is no 'Yes, occasionally' nor 'Don't know' case. 
Almost equal numbers of cases and controls were included in each sex. 

```{r}
for(ii in panels) {
  qns %>% 
    filter(RID %in% sample_id_by_panel[[ii]]) %>% 
    tabyl(Tinnitus, Sex) %>% 
    adorn_totals('col') %>% 
    kable(caption = ii) %>% 
    kable_styling(full_width = F) %>% 
    print()
}
```

#### Age distribution by sex and tinnitus status

Cases and controls were matched for age and sex. 

```{r}
qns %>% 
  ggplot() +
  aes(x = Sex, y= Age, fill= Tinnitus) +
  geom_split_violin()
```

Just the average age of females was higher than males (T-test P = `r pval_toLatex(t.test(Age ~ Sex, qns)$p.value, 3)`).

```{r}
qns %>% 
  group_by(Sex) %>% 
  summarise(
    `Avg. Age` = mean(Age),
    SEM = sd(Age) / sqrt(n())
  ) %>% 
  mutate_if(is.numeric, signif, 3) %>% 
  kable() %>% 
  kable_styling(full_width = F)
```

### Variables related to sample collection

#### Sample labs

```{r}
#  ANOVA P-value
tmp <- pval_toLatex(anova(lm(Age ~ `Sample Lab`, data = qns))[['Pr(>F)']][1], 3)
```

Samples were from `r n_distinct(qns[["Sample Lab"]])` sample labs. 
Generally speaking, the number of samples are balanced with respect to disease status and sex. 
The ages between cases and controls were well matched in each sample lab.
Note that **average ages significantly varied between the labs** (ANOVA P = `r tmp`).

```{r}
qns %>% 
  nest(data = -`Sample Lab`) %>% 
  mutate( 
    freq = map(data, . %>% tabyl(Tinnitus, Sex)),
    `P-value` = map_dbl(freq, ~ signif(fisher.test(.x)$p.value, 3)),
    `Avg. Sampling Age` = map_dbl(data, ~ signif(mean(.x$`Sampling Age`), 3))
  ) %>% 
  select(-data) %>% 
  unnest(cols= c(freq)) %>% 
  select(`Sample Lab`, Tinnitus, Male, Female, `P-value`, `Avg. Sampling Age`) %>% 
  rename_with(~ paste0(.x, footnote_marker_symbol(1)), "P-value") %>% 
  kbl(
    caption = "Age and sex of the samples from each sample lab",
    escape= F
  ) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F) %>% 
  collapse_rows(c(1, 5)) %>% 
  footnote(
    symbol = "Fisher's exact test for association between sex and Tinnitus"
  )
```

```{r}
qns %>% 
  ggplot() +
  aes(x = `Sample Lab`, y= `Sampling Age`, fill= Tinnitus) +
  geom_split_violin()
```

#### Collection date and time

The date and time of sample collection were seemingly balanced between cases and controls.

```{r, fig.height=3}
qns %>% 
  ggplot() +
  aes(x = `Collection Date`, fill= Tinnitus) +
  geom_histogram(bins= 30) +
  ggtitle('Sample collection date')

qns %>% 
  ggplot() +
  aes(x = `Collection Time`, fill= Tinnitus) +
  geom_histogram(bins= 30) +
  ggtitle('Sample collection time')
```

--------------------------------

## Proteomic data after QC

```{r}
# Olink NPX value in matrix by panel
olink_mat_by_panel <- olink %>%
  split(., .$Panel) %>%
  c(list("All panels" = olink), .) %>%
  lapply(as.matrix_olinkdf)        # to matrix
# PCA
pca <- olink_mat_by_panel %>% 
  lapply(. %>% 
           na.omit() %>% 
           prcomp(scale. = TRUE)    #  PCA
  )
```

### PCA

[Principal component analysis (PCA)](https://builtin.com/data-science/step-step-explanation-principal-component-analysis)
was applied to get a broad overview of protein profiles.
It can also identify any experimental factor that affected protein profiles extensively. 
Each dot represents a sample in the following PCA score plots. 

```{r}
# PCA
pca <- olink_mat_by_panel %>% 
  lapply(. %>% 
           na.omit() %>% 
           prcomp(scale. = TRUE)    #  PCA
  )

#' PCA plot with x/y labels including % explained
#'
#' @param prcomp_out \code{\link{prcomp}} output
#' @param mat which matrix in the \code{prcomp_out}, e.g. \code{"x"} or \code{"rotation"}
#' @param add_df data frame with additional information to visualize
#' @param id_col the column in \code{add_df} that has IDs to match
#' @param x,y components to show
#'
#' @return ggplot with labels
pca_plot <- function(prcomp_out, mat = "x", add_df = NULL, id_col = "ID", x = 1, y = 2) {
    # proportion of variance explained
    var_p <- prcomp_out$sdev^2 / sum(prcomp_out$sdev^2)
    
    ii <- paste0("PC", c(x, y))
    
    df <- prcomp_out[[mat]] %>% 
      as_tibble(rownames = "ID")
    # attach additional information data frame
    if(!is.null(add_df)) {
      stopifnot(id_col %in% names(add_df))
      df <- df %>% 
        left_join(add_df, by = c("ID" = id_col))
    }
    # prepare ggplot
    ggplot(df) +
      aes_string(x= ii[1], y= ii[2]) +
      xlab(sprintf(paste(ii[1], "(%2.1f%%)"), var_p[x] * 100)) +
      ylab(sprintf(paste(ii[2], "(%2.1f%%)"), var_p[y] * 100))
  }
```

#### By experimental factors

##### Olink analysis plate

No obvious separation by the plates was observed.

```{r, fig.cap=fig_cap}
fig_cap = paste(
  "PCR score plot.",
  "The scores of the first and second components are presented in a scatter plot,",
  "in which individual samples are colored by plate IDs.",
  "The numbers in axis labels show the proportion of variances explained by the",
  "principal components."
)
for(ii in names(pca)) {
  p <- distinct(olink, SampleID, PlateID) %>% 
    mutate(`Plate ID` = format(PlateID)) %>%  # by Plate ID
    pca_plot(pca[[ii]], add_df = ., id_col = "SampleID") +
    geom_point(aes(color = `Plate ID`)) +
    ggtitle(ii)
  print(p)
}
```

##### Sample lab

No fully separated cluster by sample lab was observed. Just some samples from Umeå (blue, N = `r sum(qns[["Sample Lab"]] == "Umeå")`) were somewhat different from the others. 

```{r, fig.cap=fig_cap}
fig_cap = paste(
  "PCR score plot.",
  "The scores of the first and second components are presented in a scatter plot,",
  "in which individual samples are colored by sample lab.",
  "The numbers in axis labels show the proportion of variances explained by the",
  "principal components."
)
for(ii in names(pca)) {
  p <- pca_plot(pca[[ii]], add_df = qns, id_col = "RID") +
    geom_point(aes(color = `Sample Lab`)) +
    ggtitle(ii)
  print(p)
}
```

#### Loadings

Loadings reflect how much each variable, here each protein, contributes to a principal component. 
Please note that only relative difference is meaningful because the sum of squares is fixed to one. 

##### Loading bi-plots

They were often visualized in a scatter plot for PC1 and PC2 as below.
The further a protein is from the origin, the more it contribute to those two principal components.

```{r, fig.cap=fig_cap}
fig_cap = paste(
  "Colors reflect the distance from the origin of each protein.",
  "The darker, the further."
)

for(ii in names(pca)) {
  p <- distinct(olink, OlinkID, Assay) %>% 
    pca_plot(pca[[ii]], "rotation", add_df = ., "OlinkID") +
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
    geom_text(aes(label = Assay, color = abs(PC1^2 + PC2^2)), size = 3) +
    scale_color_viridis_c(direction = -1) +
    theme(legend.position = "none") +
    ggtitle(ii)
  print(p)
}

```

##### Loading bar plots

The loadings were sorted and presented in bar plots.

```{r}
# Prepare loadings tibble
pca_loading <- lapply(
  pca,
  function(x) {
    x <- as_tibble(x$rotation, rownames = "OlinkID") %>% 
      right_join(distinct(olink, OlinkID, Assay, Panel), ., by = "OlinkID")
    # Fix `Assay` to handle multiple assays for the same protein
    if(anyDuplicated(x$Assay) != 0) {
      du <- unique(x$Assay[duplicated(x$Assay)]) # duplicated `Assay`
      x <- x %>% 
        mutate(
          Assay = if_else(Assay %in% du, 
                          paste0(Assay, " (", OlinkID, ")"), 
                          Assay)
        )
    }
    x %>% 
      rename(Protein = "Assay")
  }
)

# layer for the signs of loadings
layer1 <- list(
  geom_bar(stat= 'identity'),
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)),
  scale_fill_manual(
    labels= c("Positive", "Negative"),
    breaks= c("1", "-1"),
    values= c("1" = "red3", "-1" = "blue3")
  )
)
```

* 1st principal component

Basically all loadings for the 1st PC were of the same sign, positive. 
The values were decreasing gradually from highest to lowest without any greater drop. 
The PC1 might have reflected the variation primarily driven by overall magnitude of **measured** protein values. 

```{r, fig.height=4, fig.width=10}
for(ii in names(pca)) {
  p <- pca_loading[[ii]] %>% 
    arrange(desc(abs(PC1))) %>% 
    mutate(Protein = factor(Protein, levels= Protein)) %>%   # sorted by PC1
    ggplot() +
    aes(y = abs(PC1), x = Protein, fill= as.character(sign(PC1))) +
    labs(title = ii, y = "Absolute value of PC1", fill = "PC1 sign") +
    layer1
  print(p)
}
```

