---
title: "QC - clinical info table"
author: "Mun-Gwan Hong"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: 
      collapse : false
    toc_depth: 3
    # code_folding: hide
    number_sections: true
    css: styles.css
# bibliography: citations_in_report.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding= encoding, output_dir= "../reports") })
---

```{r clinic_qc-setup, echo=FALSE, include=FALSE}
## Pre-loaded packages
# Please note that many functions were called directly from the package using 
# double colon '::' to make the source clearer. 
library(tidyverse)
library(kableExtra)

## Chunk hooks
knitr::knit_hooks$set(
  optipng = knitr::hook_optipng   # Optimize PNG plots
)
## Set default chunk options
knitr::opts_chunk$set(
	echo = FALSE,    # comment out for code_folding
	# dev = c('pdf'), fig.path = '../results/figures/', # export figures to pdf files
  results = "asis",   # for HTML
  message = FALSE,    # no message from the code
  warning = FALSE,    # no warning from the code
	cache.path = "../cache/",   # cache the results from heavy analyses
  fig.width  = 8,
  fig.height = 6,
	eval.after = "fig.cap",    # to add fig. caption within the chunk
  optipng = '-o1', # '-o7 -zc9 -zm8 -zs0', # Optimize 'png' plots
	purl = FALSE      # because most of the code are not to be extracted
)
options(knitr.kable.NA = '')  # NA is shown as ''
```

```{r clinic_qc-Load data, purl=TRUE}
## File names
fn <- list(
  lib = "utils.R",
	i = list(                               # input
    c01 = "../data/s1-clinical.v01.RData"	
	)
)

# Brief check if all files exist
stopifnot(all(file.exists(unlist(fn$i), fn$lib)))

source(fn$lib)  # geom_split_violin
# Load clinical information ver.01
load(fn$i$c01)
```

* *All information was from the file, `All STOP questionnaire data 180118_v14_BloodAnalysis_v2.xlsx`*.
* *The steps that *<span style="background-color:greenyellow">modify the data (e.g. remove samples)</span> *were highlighted in the following sections.*

## Sample-wise

### Duplicated samples

Initially, there were the data of 550 patients and 550 controls in the table, `All STOP questionnaire data 180118_v14_BloodAnalysis_v2`. 
Two samples in the table were duplicated, which were both cases. 
To keep the matching condition, <span style="background-color:greenyellow">two matching controls selected by Christopher, as well as those two patients were removed</span>. 
The numbers of samples after the exclusion were `r sum(qns$Tinnitus == "Yes, always")` Tinnitus patients and `r sum(qns$Tinnitus == "No")` controls. 

There was **no duplicated sample** according to `PariticantID` after the exclusion. 

```{r}
stopifnot(anyDuplicated(qns$ParticipantID) == 0)
```

## Variable-wise

### `Age` vs. `Sampling Age`

There are two columns for age, `Sampling Age` and `Age`, which are correlated ($R^2$ = `r signif(cor(qns[["Sampling Age"]], qns$Age)^2, 3)`). 
`Age` is when tinnitus was defined, while `Sampling Age` is when blood was drawn. 

### Body-mass-index (BMI)

BMI was computed using height (`A3`) and weight (`A4`). 
`r Words(length(about_table$bmi_extreme))` extreme values were observed. 
Because they seems to be written wrong, <span style="background-color:greenyellow">the values were excluded manually</span>. 

```{r}
qns %>% 
  filter(RID %in% about_table$bmi_extreme) %>% 
  mutate(BMI = round(A4 / (A3 / 100)^2, 1)) %>% 
  select(RID, Age, Sex, BMI, `Height (cm)` = A3, `Weight (kg)` = A4) %>% 
  kbl() %>% 
  kable_styling(full_width = F)
```

### `Sample Lab`

Samples were originally collected at 4 sample labs. 
Because `MEB Plan 4` is located in Stockholm and just one female patient sample was from the lab, <span style="background-color:greenyellow">the sample was merged with the samples from `Stockholm` sample lab</span>. 

### Variables with too many missing values. 

<span style="background-color:greenyellow">`r Words(length(about_table$mostly_na))` variables having too many missing values (>80%) were removed.</span>
They are listed below.

```{r}
cat(knitr::combine_words(about_table$mostly_na, before = "`"))
```

### Variables with the same value

`r Words(nrow(about_table$one_value_only))` variables had only one identical value for all samples. 
<span style="background-color:greenyellow">Those variables without any variation were removed</span>, which are listed below.

```{r}
about_table$one_value_only %>% 
  kbl() %>% 
  kable_styling(full_width = F)
```

### Variables with one identical value or NA

`r Words(nrow(about_table$one_value_or_na))` variables had only one identical value or missing for all samples.
<span style="background-color:greenyellow">Those variables were removed.</span>

```{r}
about_table$one_value_or_na %>% 
    kable() %>% 
    kable_styling(full_width = F)
```

###  Variables with wrong code book

* `r Words(length(about_table$wrong_code$noyes))` variables for Yes / No question was supposed to be coded as `2` and `1`, according to the codebook. 
But the values found in those varaibles were `0` and `1`, <span style="background-color:greenyellow">which were translated into `No` and `Yes`, respectively</span>.

```{r}
cat(knitr::combine_words(about_table$wrong_code$noyes, before = "`"))
```

* `r Words(length(about_table$wrong_code$noyes_little_to_lot))` variables for 'No', 'Yes, a little', 'Yes, quite a lot', 'Yes, a lot' question was supposed to be coded with natural numbers from `1` to `4`, according to the codebook. 
But the values found in those variables were unexpectedly `0` to `3`, <span style="background-color:greenyellow">which were translated into 'No', 'Yes, a little', 'Yes, quite a lot', and 'Yes, a lot', respectively</span>.

```{r}
cat(knitr::combine_words(about_table$wrong_code$noyes_little_to_lot, before = "`"))
```

