---
title: "Whole samples in the study"
author: "Mun-Gwan Hong"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: 
      collapse : false
    toc_depth: 3
    # code_folding: hide
    number_sections: true
    css: styles.css
# bibliography: citations_in_report.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding= encoding, output_dir= "../reports") })
---

```{r sinfo-setup, echo=FALSE, include=FALSE}
## Pre-loaded packages
# Please note that many functions were called directly from the package using 
# double colon '::' to make the source clearer. 
library(tidyverse)
library(kableExtra)
library(janitor)

rm(list = ls())

## Chunk hooks
knitr::knit_hooks$set(
  optipng = knitr::hook_optipng   # Optimize PNG plots
)
## Set default chunk options
knitr::opts_chunk$set(
	echo = FALSE,    # comment out for code_folding
	# dev = c('pdf'), fig.path = '../results/figures/', # export figures to pdf files
  results = "asis",   # for HTML
  message = FALSE,    # no message from the code
  warning = FALSE,    # no warning from the code
	cache.path = "../cache/",   # cache the results from heavy analyses
  fig.width  = 7,
  fig.height = 5.5,
	eval.after = "fig.cap",    # to add fig. caption within the chunk
  optipng = '-o1', # '-o7 -zc9 -zm8 -zs0', # Optimize 'png' plots
	purl = FALSE      # because most of the code are not to be extracted
)
options(knitr.kable.NA = '')  # NA is shown as ''
```

```{r sinfo-Load data, purl=TRUE}
## File names
fn <- list(
  lib = "utils.R",
	i = list(                               # input
    c01 = "../data/s1-clinical.v01.RData"	
	)
)

# Brief check if all files exist
stopifnot(all(file.exists(unlist(fn$i), fn$lib)))

source(fn$lib)  # geom_split_violin, add_n
# Loading clinical information ver.01
load(fn$i$c01)
```

* *All information about samples was initially obtained from the file, `All STOP questionnaire data 180118_v14_BloodAnalysis_v2.xlsx`*.
* The results presented in this section are based on the data after QC of the given data.
**Please note that the samples _without_ proteomic data were also included during the investigation in this chapter.**
* The QC is described thoroughly in the chapter "QC of clinical data table". 

## Matching between cases and controls

### Tinnitus status and Sex

There were two groups in terms of disease status, `r knitr::combine_words(unique(qns$Tinnitus), before = "'")`. 
These is no 'Yes, occasionally' nor 'Don't know' case. Equal numbers of cases and controls were in each sex. 

```{r}
qns %>% 
  tabyl(Tinnitus, Sex) %>% 
  adorn_totals('col') %>% 
  kable() %>% 
  kable_styling(full_width = F)
```

### Age distribution by sex and tinnitus status

Cases and controls were matched for age and sex. 

```{r}
qns %>% 
  ggplot() +
  aes(x = Sex, y= Age, fill= Tinnitus) +
  geom_split_violin()
```

## Variables related to sample collection

### Sample labs

Samples were from `r n_distinct(qns[["Sample Lab"]])` sample labs. 
Generally speaking, the number of samples are balanced with respect to disease status and sex. 
The ages between cases and controls were well matched in each sample lab.
Note that **average ages significantly varied between the labs** (ANOVA P=$`r pval_toLatex(anova(lm(Age ~ `Sample Lab`, data = qns))[['Pr(>F)']][1])`$).

```{r}
qns %>% 
  nest(data = -`Sample Lab`) %>% 
  mutate( 
    freq = map(data, . %>% tabyl(Tinnitus, Sex)),
    `P-value` = map_dbl(freq, ~ signif(fisher.test(.x)$p.value, 3)),
    `Avg. Sampling Age` = map_dbl(data, ~ signif(mean(.x$`Sampling Age`), 3))
  ) %>% 
  select(-data) %>% 
  unnest(cols= c(freq)) %>% 
  select(`Sample Lab`, Tinnitus, Male, Female, `P-value`, `Avg. Sampling Age`) %>% 
  rename_with(~ paste0(.x, footnote_marker_symbol(1)), "P-value") %>% 
  kbl(
    caption = "Age and sex of the samples from each sample lab",
    escape= F
  ) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F) %>% 
  collapse_rows(c(1, 5)) %>% 
  footnote(
    symbol = "Fisher's exact test for association between sex and Tinnitus"
  )
```

```{r}
qns %>% 
  ggplot() +
  aes(x = `Sample Lab`, y= `Sampling Age`, fill= Tinnitus) +
  geom_split_violin()
```

### Collection date and time

The date and time of sample collection were seemingly balanced between cases and controls.

```{r sinfo-collection,fig.height=3}
qns %>% 
  ggplot() +
  aes(x = `Collection Date`, fill= Tinnitus) +
  geom_histogram(bins= 30) +
  ggtitle('Sample collection date')

qns %>% 
  ggplot() +
  aes(x = `Collection Time`, fill= Tinnitus) +
  geom_histogram(bins= 30) +
  ggtitle('Sample collection time')
```

## Questions possibly associated with Tinnitus

```{r}
# FUN : show contingency tables
show_tables <- function(vars_in, data_tbl) {
  for(ii in vars_in) {
    cat('*', codes$Label[codes$Variable == ii])
    # contingency table with Tinnitus status column
    sub_tbl <- data_tbl %>% 
      select(all_of(c('Tinnitus', ii))) %>% 
      drop_na() %>% 
      plyr::count(c('Tinnitus', ii)) %>% 
      pivot_wider(names_from = all_of(ii), values_from = freq, values_fill = 0)
    
    if(dim(sub_tbl)[1L] > 1 & dim(sub_tbl)[2L] > 2) {
      sub_tbl %>% 
        select(-Tinnitus) %>% 
        as.matrix() %>% 
        fisher.test(simulate.p.value = dim(sub_tbl)[2L] >= 5) %>% 
        getElement("p.value") %>% 
        pval_toLatex(3) %>% 
        cat(" (Fisher's test P = ", ., ")\n", sep = "")
    } else {
      cat("\n")
    }
    
    sub_tbl %>% 
      kable() %>% 
      kable_styling(full_width = F) %>% 
      print()
    cat("\n")
  }
}
```

#### Hearing

```{r}
# hearing related variables
i_hearing <- sapply(
  c('hearing', 'relative', 'write next to', 'hearing aids'),
  function(.x) grepl(.x, tolower(codes$Label))
) %>%
  apply(1, function(x) x[1] & all(!x[2:4])) %>% 
  {codes$Variable[.]}

show_tables(i_hearing, qns)
```

#### Smoking

```{r}
show_tables('Smoking', qns)
```

#### Stress

```{r, fig.width=6, fig.height=5}
stress <- list(
  qty = list(
    Stress = "PSQ Total score",
    Anxiety = "HADS_A Total score",
    Depression = "HADS_D Total score",
    Hyperacusis = "HQ Total score"
  ),
  cat = list(
    "Temporomandibular joint pain" = "A15_4",
    headache = "A15_1"
  )
)

qns_nm <- qns
names(qns_nm) <- make.names(names(qns_nm))

for(ii in seq_along(stress$qty)) {
  f <- formula(paste(make.names(stress$qty[[ii]]), "~ Tinnitus"))
  pval <- summary(aov(f, qns_nm))[[1]][1, 5] %>% signif(3)

  p <- ggplot(qns_nm) +
    aes_string(y = make.names(stress$qty[[ii]]), x= "Tinnitus") +
    geom_boxplot(na.rm= T) +
    ylab(stress$qty[[ii]]) +
    ggtitle(paste(names(stress$qty)[ii], "( ANOVA P =", pval, ")")) +
    add_n()
  print(p)
}
```

```{r}
show_tables(stress$cat, qns)
```

